Two applications of logic programming to Coq

### 2021-03-09 papers/types2020/revisions/conc.tex
more points in cover letter
conclusion fixed

toned down relation with eauto

We have presented two application of this synergy: one geared towards
providing a flexible approach to connecting external provers of
first-order intuitionistic logic to Coq; the other supporting an out-of-the-box way to do property-based testing for pure inductive relations.
%

@@

There are many avenues of developments for this line of research.
The code reported in Fig.~\ref{fig:augmented} is somewhat idealized in
so far that we omitted a rule performing weak-head reduction (allowing
to handle directly existential goals) and more importantly one that uses
 Elpi's \emph{delay} mechanism.\ednote{This may not be accurate anymore -am} % This a distinguished feature of Elpi, whose
% usef has been advocated in ~\cite{DunchevCT16,coen19mscs}.
The
use of constraints for data generation is well
developed~\cite{FioravantiPS15} and we could try to leverage it to
improve our PBT tactic so that it can use partially instantiated terms: for example generating
say lists with at least two elements as \texttt{[1,2|Tail]}, without
recurring to needed narrowing as in 
LazySmallCheck~\cite{smallcheck}. On the more practical side, it would be worthwhile to investigate \emph{random} generation, following the ideas in~\cite{pltredexconstraintlogic,blanco19ppdp}.

% Another current limitation of our PBT tactic is that it does not
% handle conversion within an inductive relation. One way to address
% this is to have the \emph{init} rule use Coq's unification rather than
% \lP's, via Coq-Elpi's \lsti{coq.unif-eq} hook.
%We plan to tackle this either at the Ltac level by co-routining it with simplification tactics and/or by adding rewriting features to our FPC system~\cite{ChihaniM16}.
% Mention \emph{Harpoon}, Beluga's new tactic language?

 Finally it makes sense to tie together the two threads of this paper
 and provide a way of checking and elaborating proof evidence for
 intuitionistic logic \emph{with} (inductively) defined atoms. This
 can go as far as FPC for inductive proofs~\cite{blanco15wof}.

@@

% \subsection{Future work}
% \label{ssec:fut}
%\todo{finish future work}

@@

% LocalWords:  Elpi simp LJF PBT LazySmallCheck Ltac

>>>>

We have presented two application of this synergy: one supporting an
out-of-the-box way to do property-based testing for inductive
relations; the other geared towards providing a flexible approach to
connecting external provers of first-order intuitionistic logic to Coq
%

@@

 The
code reported in Fig.~\ref{fig:augmented} is a simplification for
exposition purposes of the real implementation of the
kernel. Following ideas from bidirectional type checking, we have
factored out the product left rule in $\forall-L$ and $\supset-L$,
where the former delegates to Coq the \emph{check} that the
instantiation term $t$ is well-typed w.r.t.~$B$, while in the latter
proof search will \emph{generate} such a term, given the type
$B$. There are also other minor tweaks, such as
a rule performing weak-head reduction (allowing
us to handle directly existential goals).

There are many avenues of developments for this line of research. 
We would like to exploit one of distinguished features of Elpi: the
\emph{delay} mechanism.  The use of constraints for data generation is
well developed~\cite{FioravantiPS15} and we could try to leverage it
to improve our PBT tactic so as to generate partially instantiated
terms, % : for example generating say lists with at least two elements as
% \texttt{[1,2|Tail]}
without recurring to needed narrowing as in
LazySmallCheck~\cite{smallcheck}. On the more practical side, it would
be worthwhile to investigate \emph{random} generation, following the
ideas in~\cite{pltredexconstraintlogic,blanco19ppdp}.

Finally it makes sense to tie together the two threads of this paper
and provide a way of checking and elaborating proof evidence for
intuitionistic logic \emph{over} (inductively) defined atoms and
previously proven lemmata. This can go as far as FPC for inductive
proofs~\cite{blanco15wof}.

@@

% \subsection{Future work}
% \label{ssec:fut}
%\todo{finish future work}
% Another current limitation of our PBT tactic is that it does not
% handle conversion within an inductive relation. One way to address
% this is to have the \emph{init} rule use Coq's unification rather than
% \lP's, via Coq-Elpi's \lsti{coq.unif-eq} hook.
%We plan to tackle this either at the Ltac level by co-routining it with simplification tactics and/or by adding rewriting features to our FPC system~\cite{ChihaniM16}.
% Mention \emph{Harpoon}, Beluga's new tactic language?

@@

% LocalWords:  Elpi simp LJF PBT LazySmallCheck Ltac Coq provers FPC
% LocalWords:  lemmata

### 2021-03-09 papers/types2020/revisions/cover.tex
more points in cover letter
conclusion fixed

toned down relation with eauto

From the implementation point of view, we have extended the kernel to
account for conversion: this allows us to do prolog search and
therefore PBT on Inductive definitions that mix deduction and
computation. We have also implemented some optimizations to the kernel
itself, as briefly mentioned in the conclusion. We have experimented
with adding Hints to the kernel via Coq-Elpi's \texttt{Accumulate DB}
and while this is doable, it is not particularly useful for the
intended application, namely PBT. What we're working on is modifying
the elaboration tactic of Sec 5, which already sees the local context,
to handle Hints and Inductive. All the peices are in place, however,  given the deadline, this
will not be part of this submission.

@@

\begin{verbatim}
It is mysterious why you choose to only use constructors: you
 never justify this restriction. There is also an hint that you
 can use the Hints that eauto uses, but this is not reflected in
 your code. How can it be?
\end{verbatim}
As anticipated before, it is immediate to add another ``decide'' rule
that picks up previously proven lemmas. That would be close to the
functionality of \texttt{Hint Resolve}, but does not literally uses
Coq's Hints facility. It is feasable that to write some glue
code to connect the two, but handling all the fairly sophistcated features of the Hint
mechanism is out of the scope of this paper.

@@

We meant this paper to be a proof-of-concept rather than a system
paper. We are not in the position to carry out a ``thorough evaluation'' with
Coq's automation not with QuickChick, if not for the high-level, qualitative remarks we have made in the paper. [AM: add more if I do more about dependent terms]

@@

Well spotted: we had forgot the inequality guard in the third clause of
\texttt{insert}. The proof terms now reflects those evidence.

@@

% LocalWords:  FPC FPCs blackbox

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

>>>>

From the implementation point of view, we have extended the kernel to
account for conversion: this allows us to do Prolog search and
therefore PBT on Inductive definitions that mix deduction and
computation. We have also implemented some optimizations to the kernel
itself, as briefly mentioned in the conclusion. We have experimented
with adding Hints to the kernel via Coq-Elpi's \texttt{Accumulate DB}
and while this is doable, it is not particularly useful for the
intended application, namely PBT. What we're working on is modifying
the elaboration tactic of Sec 5, which already sees the local context,
to handle Hints and Inductive. All the pieces are in place, however,  given the deadline, this
will not be part of this submission.

@@

\begin{verbatim}
It is mysterious why you choose to only use constructors: you
 never justify this restriction. There is also an hint that you
 can use the Hints that eauto uses, but this is not reflected in
 your code. How can it be?
\end{verbatim}
As anticipated before, it is immediate to add another ``decide'' rule
that picks up previously proven lemmas. That would be close to the
functionality of \texttt{Hint Resolve}, but does not literally uses
Coq's Hints facility. It is feasible that to write some glue
code to connect the two, but handling all the fairly sophisticated features of the Hint
mechanism is out of the scope of this paper.
\begin{verbatim}
you are re-implementing a better eauto. However, your
implementation only applies as lemmas constructors of an inductive type.
Note: I understand, because I know it in advance, that a reason could be
that you plan to use LP negation as failure in the later section on testing.
Therefore you need a close world specification. If that's the main reason,
it should be made explicit in the paper.
\end{verbatim}
If we have an inductive predicate $p$ and then prove a lemma about it,
this does not violate the CWA, since the lemma is true in the same
least model of $p$.  Operationally though, it could alter NF's
capability of finding \emph{finite} failure, since the latter is
implemented by the vanilla meta-interpreter, which is not FPC-driven. At this stage
we do not see a reason to have lemmata in PBT. We have mentioned this in the paper.

@@

We meant this paper to be a proof-of-concept rather than a system
paper. We are not in the position to carry out a ``thorough
evaluation'' with Coq's automation not with QuickChick, if not for the
high-level, qualitative remarks we have made in the paper.  We have
carried out in QuickChick the example in Section 4, both with a
functional and a relational encoding. The code is present in our
examples repo. We have added a few remarks in the paper but space
prevents us form explaining the features of QuickChick and their
peculiarity. [AM: add more if I do more about dependent
terms]. Finally, we may have misstated the ambitions of our Prolog
tactic, which does not mean to compete with eauto, but merely is the
building block of FPC-driven PBT of relational specifications. We have rephrased its characterization in the paper.

@@

Well spotted: we had simply forgot the inequality guard in the third clause of
\texttt{insert}. The proof terms now reflects those evidence.

@@

% LocalWords:  FPC FPCs blackbox Coq PBT QuickChick repo CWA NF's

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
% LocalWords:  lemmata

### 2021-03-09 papers/types2020/revisions/sec3.tex
more points in cover letter
conclusion fixed

toned down relation with eauto

In this and the next section, we shall only use a much reduced subset
of the FPC framework: in essence, an FPC will be used as a simple
mechanism for bounding the search for proofs.  In our examples, a
proof certificate, denoted by the schematic variable $\XXi$, is a
particular term that is threaded throughout a logic programming
interperter.  For example, the inference rules in
Figure~\ref{fig:augmented} are augmented with an additional premise
that invokes an \emph{expert predicate} with is responsible for
extracting relevant information from a proof certificate $\XXi$ as
well as constructing continuation certificates, such as, $\XXi_1$ and
$\XXi_2$.  For example, the premise $\prodE{\XXi}{\XXi_1}{\XXi_2}{t}$
calls the expert for products which should extract from the
certificate $\XXi$ a substitution term $t$ and two continuation
certificates $\XXi_1$ and $\XXi_2$ for the two premises of this rule.
If the certificate $\XXi$ does not contain an explicit substitution
term, the expert predicate can simply return a logic variable which
would denote any term that satisfies subsequent unification problems
arising in completing the check of this certificate.

@@

The \lsti{dprolog} tactic can be seen as a programmable (and possibly
faster) version of Coq's \lsti{eauto}% , although it currently lacks the
% latter's capability of solving goals by
% \lsti{reflexivity}
; % as well as the modularity of using
% \lsti{Hints} as introduction rules
 \lsti{dprolog} is not restricted to depth-first search, since
it follows the dictates of the given FPC; for example we could easily add
iterative-deepening search. Furthermore,
%% commented out till we know for sure:
%%the tactic can still use\lsti{Hints} via Elpi's \lsti{Db} feature, and
 FPCs can provide a
\emph{trace} that may be more customizable than the one offered by 
\texttt{(e)auto}'s hard-wired \texttt{Debug} facility.

>>>>

In this and the next section, we shall only use a much reduced subset
of the FPC framework: in essence, an FPC will be used as a simple
mechanism for bounding the search for proofs.  In our examples, a
proof certificate, denoted by the schematic variable $\XXi$, is a
particular term that is threaded throughout a logic programming
interpreter.  For example, the inference rules in
Figure~\ref{fig:augmented} are augmented with an additional premise
that invokes an \emph{expert predicate} with is responsible for
extracting relevant information from a proof certificate $\XXi$ as
well as constructing continuation certificates, such as, $\XXi_1$ and
$\XXi_2$.  For example, the premise $\prodE{\XXi}{\XXi_1}{\XXi_2}{t}$
calls the expert for products which should extract from the
certificate $\XXi$ a substitution term $t$ and two continuation
certificates $\XXi_1$ and $\XXi_2$ for the two premises of this rule.
If the certificate $\XXi$ does not contain an explicit substitution
term, the expert predicate can simply return a logic variable which
would denote any term that satisfies subsequent unification problems
arising in completing the check of this certificate.

@@

The \lsti{dprolog} tactic implements some of the features of Coq's \lsti{eauto}% , although it currently lacks the
% latter's capability of solving goals by
% \lsti{reflexivity}
; % as well as the modularity of using
% \lsti{Hints} as introduction rules
 it is programmable and as such  not restricted to depth-first search, since
it follows the dictates of the given FPC; for example we could easily add
iterative-deepening search. Furthermore,
%% commented out till we know for sure:
%%the tactic can still use\lsti{Hints} via Elpi's \lsti{Db} feature, and
 FPCs can provide a
\emph{trace} that may be more customizable than the one offered by 
\texttt{(e)auto}'s hard-wired \texttt{Debug} facility.
